---
title: "Population change in Worcester"
author: "Carole Voulgaris"
date: "2023-07-21"
output: 
  rmdformats::downcute:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidycensus)
library(knitr)
```

## Population change

Did the population of Worcester County
increase a lot between 2010 and 2020? How does its increase compare
to the increase in the Boston metropolitan area (excluding New Hampshire), and
to the overall increase in the United States population?

```{r, message=FALSE}
census_vars_2020 <- c(popln = "P1_001N")
census_vars_2010 <- c(popln = "P008001")

msa_pops_2020 <- get_decennial(geography = "county",
                               state = c("MA"),
                               variables = census_vars_2020,
                               year = 2020,
                               output = "wide") |>
  filter(NAME == "Worcester County, Massachusetts" |
           NAME == "Middlesex County, Massachusetts" |
           NAME == "Essex County, Massachusetts" |
           NAME == "Suffolk County, Massachusetts" |
           NAME == "Norfolk County, Massachusetts" |
           NAME == "Plymouth County, Massachusetts") |>
  mutate(Region = ifelse(NAME == "Worcester County, Massachusetts",
                      "Worcester County", 
                      "Boston metro area (Massachusetts only)")) |>
  group_by(Region) |>
  summarise(`2020 population` = sum(popln))

msa_pops_2010 <- get_decennial(geography = "county",
                               state = c("MA"),
                               variables = census_vars_2010,
                               year = 2010,
                               output = "wide")  |>
  filter(NAME == "Worcester County, Massachusetts" |
           NAME == "Middlesex County, Massachusetts" |
           NAME == "Essex County, Massachusetts" |
           NAME == "Suffolk County, Massachusetts" |
           NAME == "Norfolk County, Massachusetts" |
           NAME == "Plymouth County, Massachusetts") |>
  mutate(Region = ifelse(NAME == "Worcester County, Massachusetts",
                      "Worcester County", 
                      "Boston metro area (Massachusetts only)")) |>
  group_by(Region) |>
  summarise(`2010 population` = sum(popln))

usa_pop_2010 <- get_decennial(geography = "us",
                         variables = census_vars_2010,
                         year = 2010,
                         output = "wide")

usa_pop_2020 <- get_decennial(geography = "us",
                         variables = census_vars_2020,
                         year = 2020,
                         output = "wide")

usa_row <- tibble(Region = "United States", 
                  `2010 population` = usa_pop_2010$popln[1],
                  `2020 population` = usa_pop_2020$popln[1])

comparison <- left_join(msa_pops_2010, msa_pops_2020) |>
  rbind(usa_row) |>
  mutate(Increase = `2020 population` - `2010 population`) |>
  mutate(`Percent increase` = paste0(formatC(100* Increase / `2010 population`, digits = 2), "%"))

kable(comparison,
      format.args = list(big.mark = ","), 
      align = c("l", "r", "r", "r", "r"))

```

## Prevalence of movers over time

One way to look at this question would be to look at how much of the 
population of the Worcester had moved between counties, but within the same state
during this period, and compare that to the population of inter-county movers
in the Boston metro area. Note that for Boston, this would include moves within
the metro area, such as moves between Suffolk county (Boston) and Middlesex county
(Cambridge).

We can get this data for both Worcester County and the Boston metro area for each year from 2010 
to 2019.

```{r, message=FALSE}
vars_acs1 <- c(moved = "B07001_049",
               total = "B07001_001")

all_years <- tibble(Region = "none",
                    move_pct = 0,
                    year = 200) |>
  filter(year ==0)

for (i in 2010:2019) {
  
  this_year <- get_acs(geography = "county",
                               state = c("MA"),
                               variables = vars_acs1,
                               year = i,
                               output = "wide",
                       survey = "acs1") |>
  filter(NAME == "Worcester County, Massachusetts" |
           NAME == "Middlesex County, Massachusetts" |
           NAME == "Essex County, Massachusetts" |
           NAME == "Suffolk County, Massachusetts" |
           NAME == "Norfolk County, Massachusetts" |
           NAME == "Plymouth County, Massachusetts") |>
  mutate(Region = ifelse(NAME == "Worcester County, Massachusetts",
                      "Worcester County", 
                      "Boston metro area (Massachusetts only)")) |>
  group_by(Region) |>
  summarise(total = sum(totalE),
            moved = sum(movedE)) |>
  mutate(move_pct = 
           moved*100/total,
         year = i) |>
    select(-total, -moved)
  
  all_years <- rbind(all_years, this_year)
  
}
```

Here's that data in a time series.

```{r}
ggplot(all_years) +
  geom_line(aes(x = year, 
                y = move_pct,
                color = Region)) +
  scale_x_continuous(breaks = seq(2010, 2019, by=1), name = "Year") +
  scale_y_continuous(name = "Percent of the population that\nmoved between counties by within\nthe same state in the past year",
                     limits = c(0, 15))
```

